\chapter{Acknowledgements}

Nothing about this thesis would have been possible without the people that have supported and put up with me over the past 4 years. If I know myself well enough, I will forget to thank someone through these acknowledgements --- however, it still is my honest sentiment that I deeply appreciate those who have been with me through this journey.

Firstly, I would like to thank my supervisor, Moritz Gerstung, for having consistently provided support and a critical and perpetually inquisitive eye to my research projects --- from his work ethic I will take with me the rigour that lies in the making of great science, as well as the taste for clear and informative communication. I also owe great appreciation to my co-supervisor, George Vassiliou, whose ambition and drive for knowledge have helped make this PhD an even more interesting project. I would also like to thank the Gerstung group for the camaraderie and for the lunches and coffee breaks-\textit{cum}-zoom meetings.

The presence of the EMBL-EBI predoc community was also instrumental to my journey and gave me a unique sense of community and unity --- it was always a helpful reminder that scientists aren't simply a bunch of nerds --- they can also be a fantastic group of people. I would like to particularly thank Conor Walker for being such a great friend over these years, for tolerating my often senseless banter and for the many fruitful beverage tastings. 

I feel like over the years I lost touch with some friends from Portugal, but the ones that remained were the best I could have wanted --- particularly Tó Zé and Cláudia, who always hosted me and kept me company over many dinners and late-night conversations. I would also like to thank Ana Rita (who was brave enough to proofread this dissertation!), Rui Pedro and Margarida for the sporadic but long conversations about any topic and for having always been there when I needed them. Some other people --- the Threshold crowd, the Ciga239 crew and João Coelho --- have also been so good at reminding me that there is more to life than what happens in science. 

I would like to profoundly thank my partner, Ashmita, for having been such a huge source of support and my unwaveringly favourite person --- you are the best part of me and I could not have possibly asked for better company in life. The distance was often tough but it was all worth it because of you.

Queria, finalmente, agradecer profundamente aos meus pais e à minha irmã e a tudo o que sempre fizeram por mim e pelo apoio incondicional, mesmo quando não lhes conseguia explicar bem o que andei a fazer durante estes 4 anos --- sem eles nunca teria chegado aqui.