\chapter{The longitudinal dynamics and natural history of clonal haematopoiesis}

\section{Introduction}

\ac{ch} has been widely studied regarding its likely associations with specific blood indices, smoking history, chemotherapy, inflammation and age \cite{Dawoud2020-af,Coombs2017-ph,McKerrell2015-rl,Cai2018-yi,Bolton2020-ct}, and has been shown to be associated with an increased risk of cardiovascular disease and haematological cancer \cite{Jaiswal2014-rl,Genovese2014-eu, Young2016-du, Xie2014-np, Desai2018-pj, Midic2020-zh, Zink2017-zi, Acuna-Hidalgo2017-ng}. However, the study of the dynamics and evolution of \ac{ch} is still in its infancy - indeed, while it is now known that fitness advantages are gene- and site-dependent \cite{Watson2020-pz,Robertson2021-sw}, previous work on this topic have only been able to clarify either the average behaviour of clones or the clonal dynamics at an older age. While both sound relatively complimentary, apparent inconsistencies have arisen. Indeed, while \cite{Watson2020-pz} used the distribution of clone sizes to infer that \ac{dnmt3a} mutations can lead to some of the fastest growing clones, the work in \cite{Robertson2021-sw}, which uses longitudinal targeted sequencing, reaches different conclusions: \ac{dnmt3a} actually leads to some of the slowest growing clones in old age, a period during which mutations in splicing genes (particularly in \ac{u2af1} and \ac{sf3b1}). As such, there seems to be an apparent paradox - while the distribution of clone sizes would imply that \ac{dnmt3a} clones expand quickly over life, these fast dynamics are not observable during old age. 

The work here presented contributes to this body of knowledge and, to some extent, resolves the aforementioned contradiction. Here, I have analysed data from longitudinal targeted sequencing for over 300 older individuals and confirm several of the findings presented by the previous by Robertson \textit{et al.} using longitudinal sequencing \cite{Robertson2021-sw}. Additionally, we are able to disentangle clone-specific (or "unknown-cause") effects from those attributable to specific genes and sites. This revealed that mutations in splicing genes, particularly in \ac{u2af1} and \ac{srsf2}-P95H, confer the greatest fitness advantage in older age, while mutations in \ac{dnmt3a} conferred a relatively small fitness advantage, similarly to \cite{Robertson2021-sw}. To complement this evolutionary snapshot of clonal dynamics during older age, I also used phylogenetic data from single-cell colonies derived for three of the individuals for which we had longitudinal targeted sequencing data and from four additional older individuals from a publication by Michell \textit{et al.} \cite{Mitchell2021-zl} and infer the phylodynamic trajectories for all expansions, with and without known drivers. Together with estimates from the longitudinal targeted sequencing data, this helped us clarify the apparent contradiction between \cite{Robertson2021-sw} and \cite{Watson2020-pz} - most clones decelerate through life, an effect that is likely to be motivated by competition driven by an oligoclonal \ac{hsc} population and affects lesser fit clones the most. Additionally, I use the fitness estimates from longitudinal targeted sequencing to establish associations between both selection in \ac{aml} and \ac{mds} and risk of \ac{aml} onset as calculated in \cite{Abelson2018-wh}.

\subsection{Contribution and disclaimer}

The work presented in this chapter is also available as a preprint in \cite{Fabre2021-uw} and some parts may coincide. My contribution to this work, consisted on the analysis of longitudinal data and generation of all figures associated with it, as well as all the analyses associated with phylodynamic estimation of growth coefficients and phylogenetic estimation of ages at onset.

\section{Methodology}

\subsection{Cohort composition}

I worked exclusively with variant calls and, as such, will not go into detail about the variant calling process; nonetheless, for reference, these are available in . In regards to the cohort here studied, the data I used consisted of sequencing data for the 17 aforementioned genes across 385 individuals. Notably, 80 (20.8\%) of these individuals had no mutations on any of the genes under study and were disregarded. For each individual, blood was collected at between 2 and 5 timepoints (Table \ref{table:age-dist}) across a median period of 12.9 years (3.21-16.1) and 5 phases (Phase 1-5), with 96.4\% of all individuals having more than two timepoints. Each blood sample was sequenced for a panel of 56 genes at a median coverage of 1350X. Only 17 of these genes - as represented in \ref{table:genes-sites-modelling} - were used after assessing which ones were under positive selection \cite{Fabre2021-uw,Martincorena2017-ii}. For 13 individuals blood was collected at an additional phase - Phase 6 - for validation and for three of these individuals (IDs 2259, 3877 and 500) at timepoint 6 single-cell colonies were derived and sequenced (85 colonies for 2259, 93 colonies for 3877 and 92 colonies for 500). This latter process was used to build single-cell phylogenies, a process that will be detailed further ahead. The specific selection of these individuals was informed in a particular way - mutations in the spliceosome, due to their age-dependency, are of high interest and motivated the derivation and sequencing of 96 single-cell colonies for three individuals with large (>5\% VAF) spliceosome mutations. Particularly, the mutations of interest were \ac{sf3b1}-K666N, \ac{u2af1}-Q157R and \ac{srsf2}-P95H.

\begin{table}
\centering
\caption{Distribution of individuals with a given number of timepoints.}
\pgfplotstabletypeset[
string type,
columns/ntp/.style={
    column name=Number of timepoints,
    column type={C{.2\textwidth}}},
columns/n/.style={
    column name=Number of individuals,
    column type={C{.2\textwidth}}},
columns/p/.style={
    column name=Proportion,
    postproc cell content/.append code={\pgfkeysalso{@cell content/.add={}{\%}}},
    column type={C{.2\textwidth}}},
columns/cp/.style={
    column name=Cumulative proportion,
    postproc cell content/.append code={\pgfkeysalso{@cell content/.add={}{\%}}},
    column type={C{.3\textwidth}}},
every head row/.style={before row={\toprule},after row=\midrule},
every last row/.style={after row={\toprule}},
every odd row/.style={before row={\rowcolor[gray]{0.9}}}
]\ageDist
\label{table:age-dist}
\end{table}

\subsection{Modelling clonal trajectories from longitudinal targeted sequencing data}

I assume that the \ac{vaf} at each timepoint is a proxy for clone size in the bone marrow and assume that each mutation is monoclonal. Both these assumptions are supported by previous studies \cite{Miles2020-fz,Lee2020-yp,Hwang2018-tp} and allow us to model each clonal trajectory --- the collection of \ac{vaf} values for a single driver mutation in a single individual --- as a time-dependent scaled sigmoid with an asymptotic value at 0.5. The reason why 0.5, rather than 1, is used is to accommodate the fact that most mutations are likely to be haploid. While phasing these mutations would be the ideal case scenario this is not possible using this targeted sequencing panel.

As mentioned, each clonal trajectory is characterized by a time-dependent sigmoid curve. To parametrize it, I have, for mutation $j$ in individual $i$, four distinct parameters: $b_{gene_j}$ characterizing the fitness advantage conferred by gene identity, $b_{site_j}$ characterizing the fitness advantage conferred by site identity for non-truncating mutations in sites occurring in two or more individuals, $b_{unknown_{ij}}$ characterizing the fraction of growth that depends on clone-specific factors and $u_{ij}$ which is an offset for the sigmoid curve to account for clones appearing at different times. As such, the expected proportion of mutant counts on site $i$ in individual $j$ at time $t$ is given by the following model: $p_{ij}(t)=sigmoid((b_{gene_j}+b_{site_j}+b_{unknown_{ij}})*t + u_{ij})*0.5$. In our case and to speed up convergence, $t$ is calculated as the age of each individual after subtracting the youngest age in the cohort. With effect, the hierarchical model used here is quite similar to the one used in the previous chapter, with the main exception being the decomposition of $b_{mut_j}$ in the previous chapter into $b_{gene_j}$ and $b_{site_j}$ when a site is present in two or more individuals or the replacement of $b_{mut_j}$ with $b_{gene_j}$ when the mutation is present in a single individual.

Using this model I fit each parameter through a \ac{mcmc} approach. I assume as priors for $b_{gene_j}$ and $b_{site_j}$ a normal distribution with mean 0 and standard deviation 0.1, for $b_{unknown_{ij}}$ a normal distribution with mean 0 and standard deviation 0.05 and for $u_{ij}$ a uniform distribution between -50 and 0. The prior for $\beta$ follows a truncated normal distribution whose parameters were inferred as described in the previous chapter. For the sampling, I use \ac{hmc} as implemented in greta \cite{Golding2018-zp} with a number of leapfrog steps between 150 and 300. I draw 5000 samples and discard the initial 2500 samples, thus estimating the mean, median, credible intervals and \ac{hpdi} of each parameter using the last 2500 samples. The quality control step described in the previous section - calculating the likelihood of each sample under our model and discarding all trajectories with at least one point whose likelihood has a tail probability of over $97.5\%$. I calculate the age at onset as described in the previous chapter - $t_{onset,ij} = \frac{log(g/s_{ij}/N)-1-s_{ij}}{s_{ij}}$.

\subsubsection{Preventing identifiability issues and reducing uninformed estimates.}

To address possible identifiability issues in this model, when a gene has a single mutation (\ac{jak2} with V617F and \ac{idh2} with R140Q), the effect is considered to occur at the site level. To avoid estimating the dynamics of a site from a single individual, I only model sites explicitly when two or more individuals have a missense mutation on that site - I refer to these sites as “recurrent sites”. For all other sites where a single mutation is observed across the entire cohort their differential dynamics will be contained in the unknown cause effect. Ultimately, I consider a total of 17 genes and 39 recurrent sites (Table \ref{table:genes-sites-modelling}).

\begin{table}
\centering
\caption{Summary of different effects in the model. “X” indicates the utilization of that specific effect (Truncating and Non-truncating) for each gene and in Sites I detail the sites that were explicitly modelled.}
\pgfplotstabletypeset[
string type,
columns/g/.style={
    column name=Gene,
    postproc cell content/.style={@cell content=\textit{##1}},
    column type={C{.1\textwidth}}},
columns/t/.style={
    column name=Truncating effect,
    column type={C{.20\textwidth}}},
columns/nt/.style={
    column name=Non-truncating effect,
    column type={C{.20\textwidth}}},
columns/sites/.style={
    column name=Explicitly modelled sites (absolute prevalence),
    column type={C{.4\textwidth}}},
every head row/.style={before row={\toprule},after row=\midrule},
every last row/.style={after row={\toprule}},
every odd row/.style={before row={\rowcolor[gray]{0.9}}}
]\genesSites
\label{table:genes-sites-modelling}
\end{table}

\subsubsection{Assessing the predictive performance of clonal growth predictions}
Using an additional time-point (phase 6) available for 11 individuals with mutations in \ac{cbl} (c.2434+1G>A), \ac{dnmt3a} (P385fs, R882H, W330X), \ac{gnb1} (K57E), \ac{jak2} (V617F), \ac{ppm1d} (Q524X), \ac{SF3B1} (K666N, K700E, R625L), \ac{SRSF2} (P95H, P95L), \ac{tet2} (Q1542X) and \ac{u2af1} (Q157P, Q157R). Using the model described in this section of the Methods and conditioning on the previous timepoints, we predict the additional time-point and assess the predictive performance through the \ac{mae} to the true \ac{vaf} value.

\subsection{Analysing the impact of smoking history, sex, age and smoking experience on the clonal growth rate}

The association between unknown-cause growth and age was calculated using the Pearson correlation considering all genes, both together and separately while controlling for multiple testing. For the association between unknown-cause growth and sex and smoking history, a multivariate regression was used, with unknown-cause growth as the dependent variable and sex and previous smoking experience as the covariates while controlling for age.

\subsection{Phylogenetic and phylodynamic estimation of clonal trajectories from single-cell derived phylogenetic trees}

Single-cell colonies were derived and sequenced for three individuals and their phylogenetic trees inferred as detailed in \cite{Fabre2021-uw}. Each tree was made ultrametric by iteratively extending branches until the distance between any two tips is no bigger than the distance between either tip and any third tip (also known as the three point condition) and scaled in such a way that they span the age of the individual, with an 0.75 years accounting for gestational development. In these trees, we define and detect expansions as either clades with one known driver mutation at their \ac{mrca} or as clades whose \ac{mrca} branch length is at least 10\% of the depth of the tree. This allows us to detect expansions with no known driver. The determination of clonal trajectories from each expansion, as well as the quantification of growth from these trajectories, is described in the previous chapter.

The motivation to scale the trees according to the age of the individual plus 9 months (0.75 years) comes from recent work by \etal{Abascal} showing that mutation rates are constant through life \cite{Abascal_2021_gjvqfm} and, as such, mutations should accumulate linearly with age --- while it is possible that there is a short period of greater mutation accumulation before birth \cite{SpencerChapman_2021_gjz4x4}, the amount of error incurred by assuming that mutation rates remains constant through life and gestational development is relatively small. 

\ac{Detecting deceleration in single-cell phylogenies and longitudinal data}

We infer the presence of deceleration in both single-cell phylogenies and longitudinal data. To do this, we use two distinct methods: calculating the ratio between the expected and observed \ac{vaf} and calculating deceleration using growth rates. 

For the first method --- calculating the ratio between expected and observed \ac{vaf} - we use the value for the early growth from the biphasic log-linear fit described in the previous chapter and extrapolate the \ac{eps} to the age at sampling. By doing so we get the expected clone fraction if growth had not changed during the \ac{eps} trajectory. We also calculated the observed clone fraction as the fraction of tips in the clade. To get the expected clone fraction from \ac{eps} we divide \ac{eps} by the inferred population size in \etal{Lee-Six} (200,000 HSC) \cite{Lee-Six2018-lp}. We then calculate the ratio between the expected and observed clone size --- if this ratio is close to 1 this implies little to no changes in dynamics, whereas a ratio above 1 implies deceleration and a ratio below 1 implies acceleration.

For the second method --- calculating deceleration using growth rates --- we define two distinct quantities for both single-cell phylogenies/longitudinal data --- expected/observed growth, corresponding to the growth rate of each clone during observation at old age, and early/minimal historical growth, corresponding to the growth rate of each clone at an earlier stage of clonal dynamics --- and calculate the ratio between them.

As such, for phylogenies we first calculate the \ac{eps} trajectory for each clade using \ac{bnpr}. Next, and using their \ac{eps} trajectory, we calculate their expected growth rate by assuming a sigmoid growth. We additionally assume that the final \ac{eps} (\ac{eps} at sampling) estimate corresponds to the fraction of tips in the clade and we scale our data accordingly such that the unit corresponds to the maximum \ac{eps} and the fraction of tips in the clade corresponds to \ac{eps} at sampling. Thirdly and using the biphasic log-linear fit described in “Validating annual growth rate inferences from single-cell phylogenies with Wright-Fisher simulations” we derive the value for early growth. Finally, as a measure of deceleration, we calculate the ratio between expected and early growth --- a value close to 1 for this ratio implies an absence of deceleration whereas smaller values imply deceleration. 

For the longitudinal data we use the observed growth for each clone, calculated as described above. Next, we calculate the (minimal) historical growth as the growth that excludes all posterior samples that would lead to age at onset estimates exceeding lifetime (ages at onset for clones appearing below the age of -1, a heuristic value chosen to represent developmental onset of clones). Finally and as a measure of deceleration, we calculate the ratio between observed and historical growth. The interpretation for this ratio is similar to that defined in the previous paragraph for phylogenetic data --- a value of 1 implies an absence of detectable deceleration, whereas smaller values represent the minimal amount of deceleration. This method has, however a caveat --- due to the nature of this calculation (excluding posterior samples which are too slow to provide solutions within lifetime), values above 1 (indicating acceleration) are technically impossible. 

\section{Results}

\section{Discussion}

\section{Code availability}

The code for this analysis and the associated notebooks are available in \url{https://gerstung-lab.github.io/ch-dynamics/}, more specifically in "Growth rate coefficients and age at onset inference, possible associations with phenotype", "Analysis of the phylogenetic trees in Mitchell et al. (2021)" and "Investigating the historical growth effect and poor fits".